---
title: "virtual_IA"
author: "Anjolie"
date: "2/9/2022"
output: 
  html_document:
    number_sections: true
    toc: true
---

```{r setup, include=FALSE}
##add table of contents
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
require(knitr)
require(mosaic)
require(lattice)
require(ggplot2)
require(tigerstats)
require(forcats)
require(dplyr)
library(magrittr)

#raw
virtual_raw <- read_csv("virtual_raw.csv", 
    col_types = cols(
      ValueType = col_skip(), 
        `Location type` = col_skip(), `Period type` = col_skip(), 
        `Dim1 type` = col_skip(), `Dim2 type` = col_skip(), 
        `Dim3 type` = col_skip(), Dim3 = col_skip(), 
        Dim3ValueCode = col_skip(), DataSourceDimValueCode = col_skip(), 
        DataSource = col_skip(), FactValueNumericPrefix = col_skip(), 
        FactValueUoM = col_skip(), FactValueNumericLowPrefix = col_skip(), 
        FactValueNumericHighPrefix = col_skip(), 
        FactValueTranslationID = col_skip(), 
        FactComments = col_skip(), 
      IsLatestYear = col_skip(), Indicator = col_skip(), IndicatorCode = col_skip(), ParentLocationCode = col_skip(), ParentLocation = col_skip(), Dim2ValueCode = col_skip(), Language = col_skip(), DateModified = col_skip()
        )
    )
##location as factor
virtual_raw$location <- as.factor(virtual_raw$Location)
virtual_raw <- select(virtual_raw, -Location)

##disease as factor
virtual_raw$disease <- as.factor(virtual_raw$Dim2)
virtual_raw <- select(virtual_raw, -Dim2)

##sex as a factor
virtual_raw$sexCat <- as.factor(virtual_raw$Dim1)
virtual_raw <- select(virtual_raw, -Dim1)

##new factor for 'margin of error' (is it a margin? check WHO website for how highs and lows were calculated)


# MAKING SUBSETS
##by country
raw_JP <- subset(virtual_raw, location == "Japan")
  raw_JP <- select(raw_JP, -location)
raw_US <- subset(virtual_raw, virtual_raw$location =="United States of America")
  raw_US <- select(raw_US, -location)
raw_UK <- subset(virtual_raw, virtual_raw$location == "United Kingdom of Great Britain and Northern Ireland")
  raw_UK <- select(raw_UK, -location)

#raw, only diabetes and resp diseases
rawUS_diabetes <- subset(raw_US, raw_US$disease =="Diabetes mellitus")
rawUS_resp <- subset(raw_US, raw_US$disease =="Respiratory diseases")
## for Japan
rawJP_diabetes <- subset(raw_JP, raw_JP$disease =="Diabetes mellitus")
rawJP_resp <- subset(raw_JP, raw_JP$disease =="Respiratory diseases")
## UK
rawUK_diabetes <- subset(raw_UK, raw_UK$disease =="Diabetes mellitus")
rawUK_resp <- subset(raw_UK, raw_UK$disease =="Respiratory diseases")

#US by sex categories
rawUS_allSex <- subset(raw_US, raw_US$sexCat == "Both sexes")

rawUSdiabetes_allSex <- subset(rawUS_diabetes, rawUS_diabetes$sexCat =="Both sexes")
  rawUSdiabetes_2sex <- rawUS_diabetes[!(rawUS_diabetes$sexCat=="Both sexes"),]

  #female vs male
    rawUSdiabetes_F <- subset(rawUS_diabetes, rawUS_diabetes$sexCat =="Female")
    rawUSdiabetes_M <- subset(rawUS_diabetes, rawUS_diabetes$sexCat =="Male")

rawUSresp_allSex <- subset(rawUS_resp, rawUS_resp$sexCat =="Both sexes")
  rawUSresp_2sex <- rawUS_resp[!(rawUS_resp$sexCat=="Both sexes"),]
   
   #female vs male
  rawUSresp_F <- subset(rawUS_resp, rawUS_resp$sexCat =="Female")
  rawUSresp_M <- subset(rawUS_resp, rawUS_resp$sexCat =="Male")

  
##JP by sex cats
rawJP_allSex <- subset(raw_JP, raw_JP$sexCat == "Both sexes")

rawJPdiabetes_allSex <- subset(rawJP_diabetes, rawJP_diabetes$sexCat =="Both sexes")
  rawJPdiabetes_2sex <- rawJP_diabetes[!(rawJP_diabetes$sexCat=="Both sexes"),]

rawJPresp_allSex <- subset(rawJP_resp, rawJP_resp$sexCat =="Both sexes")
  rawJPresp_2sex <- rawJP_resp[!(rawJP_resp$sexCat=="Both sexes"),]
  
   #female vs male, diab
    rawJPdiabetes_F <- subset(rawJP_diabetes, rawJP_diabetes$sexCat =="Female")
    rawJPdiabetes_M <- subset(rawJP_diabetes, rawJP_diabetes$sexCat =="Male")
   
   #female vs male, resp
  rawJPresp_F <- subset(rawJP_resp, rawJP_resp$sexCat =="Female")
  rawJPresp_M <- subset(rawJP_resp, rawJP_resp$sexCat =="Male")
  
##UK by sex cats

rawUK_allSex <- subset(raw_UK, raw_UK$sexCat == "Both sexes")

rawUKdiabetes_allSex <- subset(rawUK_diabetes, rawUK_diabetes$sexCat =="Both sexes")
  rawUKdiabetes_2sex <- rawUK_diabetes[!(rawUK_diabetes$sexCat=="Both sexes"),]

rawUKresp_allSex <- subset(rawUK_resp, rawUK_resp$sexCat =="Both sexes")
  rawUKresp_2sex <- rawUK_resp[!(rawUK_resp$sexCat=="Both sexes"),]
  
  #female vs male, diab
    rawUKdiabetes_F <- subset(rawUK_diabetes, rawUK_diabetes$sexCat =="Female")
    rawUKdiabetes_M <- subset(rawUK_diabetes, rawUK_diabetes$sexCat =="Male")
   
   #female vs male, resp
  rawUKresp_F <- subset(rawUK_resp, rawUK_resp$sexCat =="Female")
  rawUKresp_M <- subset(rawUK_resp, rawUK_resp$sexCat =="Male")
  

  
if(!require("rglwidget")) {install.packages("rglwidget");require("rglwidget")}
```

# title, exploration, etc

$H_A$: In the US, there's a significant difference in the number of deaths from diabetes between the sexes. 

$H_0$: In the US, the number of deaths from diabetes is independent from sex. 

# US bar chart, between sexes 

```{r}
#diabetes
##setting percents manually? doesn't need an extra graph. just find ratio
resF <- c(sum(rawUSdiabetes_F$FactValueNumeric))
resM <- c(sum(rawUSdiabetes_M$FactValueNumeric))
resAll <- c(sum(rawUSdiabetes_allSex$FactValueNumeric))

basicCN <- ggplot(rawUSdiabetes_2sex,aes(x=sexCat,y=FactValueNumeric, fill= sexCat))
basicCN + geom_boxplot(notch= FALSE ,outlier.color = "red") + stat_boxplot(geom ='errorbar') + 
  labs(title = expression (atop(bold("Diabetes in the US by sex (total deaths)"), 
  atop(italic("Females: 48.8%, Males: 51.2%")))), x = "sex", y= "deaths") +
scale_fill_discrete(name= "Sex")

F_perc <- resF/resAll
  print(F_perc)
M_perc <- resM/resAll
  print(M_perc)
```

  The median deaths, as predicted, were around 36,700 for each sex. But the whiskers (margin of error) show a skew towards males; almost 50K men could have died, while the upper whisker for women is lower than 40K. 
  
```{r sex_summaries, include=FALSE}
#numerical  
#setup (making models for regression). mention y= mx+b equations in analysis. 
  model.USdiabetes_F <- lm(FactValueNumeric ~ Period, data = rawUSdiabetes_F)
  model.USdiabetes_M <- lm(FactValueNumeric ~ Period, data = rawUSdiabetes_M)
  
  #cleaning up (fem summary)
USdiabetes_F <- select(rawUSdiabetes_F, -SpatialDimValueCode)
USdiabetes_F <- select(USdiabetes_F, -Dim1ValueCode)
USdiabetes_F <- select(USdiabetes_F, -Value)
USdiabetes_F <- select(USdiabetes_F, -sexCat)
USdiabetes_F <- select(USdiabetes_F, -disease)
#separate summary df to remove period from summ but not graphs
USdiabetes_FSummary <- select(USdiabetes_F, -Period)
#rename main value
USdiabetes_FSummary <- USdiabetes_FSummary %>%
rename(ProbableDeaths = FactValueNumeric)

#cleaning up (male summary)
USdiabetes_M <- select(rawUSdiabetes_M, -SpatialDimValueCode)
USdiabetes_M <- select(USdiabetes_M, -Dim1ValueCode)
USdiabetes_M <- select(USdiabetes_M, -Value)
USdiabetes_M <- select(USdiabetes_M, -sexCat)
USdiabetes_M <- select(USdiabetes_M, -disease)
#separate summary df to remove period from summ but not graphs
USdiabetes_MSummary <- select(USdiabetes_M, -Period)
#rename main value
USdiabetes_MSummary <- USdiabetes_MSummary %>%
rename(ProbableDeaths = FactValueNumeric)

```

## F numerical (for boxplot)

```{r}
  summary(USdiabetes_FSummary)
```
The boxplot only shows the probable deaths. Diabetes would kill an If the null hypothesis is true, about 35,000 deaths per sex would be expected annually. 
  

## M numerical (for boxplot)

```{r}
  summary(USdiabetes_MSummary)
```

]]average deaths, each gender

# US diabetes line graph (regardless of sex)


```{r summary}
require(scales)
#graphical model w/ regression. For US diabetes regardless of sex. 
model.USdiabetesAllsex <- lm(FactValueNumeric ~ Period, data = USdiabetes_allSex)
pred1 <- predict(model.USdiabetesAllsex, interval="prediction")
#model for factvalue Highs and Lows
model.USdiabHigh <- lm(FactValueNumericHigh ~ Period, data = USdiabetes_allSex)
model.USdiabLow <- lm(FactValueNumericLow ~ Period, data = USdiabetes_allSex)

#actual graph
new_USdiabetesAllsex <- cbind(USdiabetes_allSex, pred1)
ggplot(new_USdiabetesAllsex, aes(Period, FactValueNumeric))+
    geom_point() + expand_limits(y= 55000) +
    geom_line(aes (Period, FactValueNumericHigh), color = "red", linetype = "dashed") +
    geom_line(aes (Period, FactValueNumericLow), color = "red", linetype = "dashed") +
    geom_smooth(method=lm, se=TRUE) + labs(title = "US diabetes (all sexes) over time", x= "year", y = "deaths") + scale_y_continuous(labels = comma)
```
 
  
### Numerical - regression summary
```{r}
summary.lm(model.USdiabetesAllsex)
```
The regression line (blue on the graph) is [deaths] = 677.1 * [year] - 1284550.5. 

## Numerical - general summary 

```{r, include=FALSE}
#graph setup
USdiabetes_allSex <- select(rawUSdiabetes_allSex, -Dim1ValueCode)
USdiabetes_allSex <- select(USdiabetes_allSex, -Value)
USdiabetes_allSex <- select(USdiabetes_allSex, -sexCat)
USdiabetes_allSex <- select(USdiabetes_allSex, -disease)
USdiabetes_allSex <- select(USdiabetes_allSex, -SpatialDimValueCode)

##remove period later so the graph still works
USdiabetes_allSexSummary <- select(USdiabetes_allSex, -Period)

USdiabetes_allSexSummary <- USdiabetes_allSexSummary %>%
rename(ProbableDeaths = FactValueNumeric)
```

```{r}
summary(USdiabetes_allSexSummary)
```

The mean annual death count is 75,993, so if the null hypothesis was true then an average of 38K of each sex would die annually. 


# by sex (line graph)

```{r}
#female vs male, diabetes

  #diabetes
plot(rawUSdiabetes_M$Period, rawUSdiabetes_M$FactValueNumeric, type = "l",  col="blue",                               
     main = "US diabetes by sex, over time",
     xlab = "year",
     ylab = "deaths")
lines(rawUSdiabetes_F$Period, rawUSdiabetes_F$FactValueNumeric, type = "l", col = "red") 
    #ablines for male, fem
abline(reg = lm(rawUSdiabetes_M$FactValueNumeric ~ rawUSdiabetes_M$Period), col = "blue", lty = 2)
  abline(reg = lm(rawUSdiabetes_F$FactValueNumeric ~ rawUSdiabetes_F$Period), col = "red", lty = 2)
    #legend
legend("topleft",                                       # Add legend to plot
       legend = c("Female", "Male"),
       col = c("red", "blue"),
       lty = 1)
legend("top", 
       legend = c("F (best fit)", "M (best fit)"), 
       col = c("red", "blue"), 
       lty = 2)
```
  
  Diabetes killing more females before 2010 might be explained by men dying earlier from other dangers like workplace hazards; modern society, due to expanded safety regulation in the workplace and otherwise, is much safer. But the upward spike afterwards (in male diabetes deaths) is still suspect and probably has other explanations, like the 2008 recession which introduced compounding dangers like starvation due to a spike of mass unemployment. Even with the slow drop-off in unemployment after 2010, the recession seems to be have long-lasting effects. Overall, males are killed far more often by diabetes than females. Diabetes deaths had a small spike in 2005. 
  
### regression summaries
```{r}
#setup (making models for regression). mention y= mx+b equations in analysis. 
  model.USdiabetes_F <- lm(FactValueNumeric ~ Period, data = rawUSdiabetes_F)
  model.USdiabetes_M <- lm(FactValueNumeric ~ Period, data = rawUSdiabetes_M)
  
#female
summary.lm(model.USdiabetes_F)
#male
summary.lm(model.USdiabetes_F)
```

# chi test for boxplot(s)

```{r}
#if numCat and other sht doesn't work, copy those RMDs from a WUP
```

# two way anova on female/male regressions over time (x= period)

```{r}
# Fem vs male
# fem vs allsex
# male vs allsex
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
